---
title: "DSCI 401 Project"
author: "Rachel Jordan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#load data
crashes <- read.csv("Crashes.csv")
people <- read.csv("People.csv")
vehicles <- read.csv("Vehicles.csv")

#load tidyverse
require(tidyverse)

#pare down variables to only the ones I want
crashes <- crashes %>% 
  select(CRASH_RECORD_ID,CRASH_DATE,POSTED_SPEED_LIMIT,WEATHER_CONDITION,
         LIGHTING_CONDITION,CRASH_TYPE,DAMAGE,PRIM_CONTRIBUTORY_CAUSE,
         SEC_CONTRIBUTORY_CAUSE,STREET_NO,STREET_DIRECTION,STREET_NAME,
         MOST_SEVERE_INJURY,INJURIES_TOTAL,INJURIES_FATAL,
         INJURIES_INCAPACITATING,INJURIES_NON_INCAPACITATING,
         INJURIES_REPORTED_NOT_EVIDENT, INJURIES_NO_INDICATION,
         INJURIES_UNKNOWN,CRASH_HOUR,CRASH_DAY_OF_WEEK,CRASH_MONTH,
         LATITUDE,LONGITUDE,LOCATION)

people <- people %>% 
  select(PERSON_ID,PERSON_TYPE,CRASH_RECORD_ID,SEX,AGE,DRIVERS_LICENSE_STATE,
         EJECTION,INJURY_CLASSIFICATION,DRIVER_ACTION,DRIVER_VISION,
         BAC_RESULT,BAC_RESULT.VALUE,CELL_PHONE_USE)

vehicles <- vehicles %>% 
  select(CRASH_RECORD_ID,MAKE,MODEL,LIC_PLATE_STATE,VEHICLE_YEAR)

#join data sets
crashes_and_people <- people %>% 
  left_join(crashes,by=c("CRASH_RECORD_ID" = "CRASH_RECORD_ID"))
#each record is a person

crashes_and_vehicles <- vehicles %>% 
  left_join(crashes,by=c("CRASH_RECORD_ID" = "CRASH_RECORD_ID"))
#each record is a vehicle

```

#### Profile of typical drivers at each crash severity

```{r}
#filter for only driver information
drivers <- crashes_and_people %>% 
  filter(PERSON_TYPE == "DRIVER")

#filter out blank gender (97 records are blank for gender)
drivers <- drivers %>% 
  filter(SEX == "F" | SEX == "M" | SEX == "X")
#unclear if X is non-binary, NA, or a mix

#change drivers license state to local or other
drivers <- drivers %>% 
  drop_na(DRIVERS_LICENSE_STATE) %>% 
  mutate(WHERE_FROM = case_when(
    DRIVERS_LICENSE_STATE == "IL" ~ "Local",
    TRUE ~ "Other"
  ))

#find profile of most frequent drivers in fatal crashes
fatal_drivers <- drivers %>% 
  filter(MOST_SEVERE_INJURY == "FATAL")

fatal_driver_profile <- fatal_drivers %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(fatal_drivers)) 

fatal_driver_profile[which.max(fatal_driver_profile$proportion),] #male locals make up 53.6% of fatal crashes in Chicago

#find profile of most frequent drivers in incapacitating crashes
inc_drivers <- drivers %>% 
  filter(MOST_SEVERE_INJURY == "INCAPACITATING INJURY")

inc_driver_profile <- inc_drivers %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(inc_drivers)) 

inc_driver_profile[which.max(inc_driver_profile$proportion),] #male locals make up 47.2% of incapacitating crashes in Chicago

#find profile of most frequent drivers in non-incapacitating injury crashes
noninc_drivers <- drivers %>% 
  filter(MOST_SEVERE_INJURY == "NONINCAPACITATING INJURY")

noninc_driver_profile <- noninc_drivers %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(noninc_drivers)) 

noninc_driver_profile[which.max(noninc_driver_profile$proportion),] #male locals make up 44.7% of non-incapacitating injury crashes in Chicago

#find profile of most frequent drivers in "injuries reported but not evident" crashes
ne_drivers <- drivers %>% 
  filter(MOST_SEVERE_INJURY == "REPORTED, NOT EVIDENT")

ne_driver_profile <- ne_drivers %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(ne_drivers)) 

ne_driver_profile[which.max(ne_driver_profile$proportion),] #male locals make up 43.1% of "injuries reported but not evident" crashes in Chicago

#find profile of most frequent drivers in "no injury" crashes
noinj_drivers <- drivers %>% 
  filter(MOST_SEVERE_INJURY == "NO INDICATION OF INJURY")

noinj_driver_profile <- noinj_drivers %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(noinj_drivers)) 

noinj_driver_profile[which.max(noinj_driver_profile$proportion),] #male locals make up 40.6% of crashes in Chicago with no injuries
```

#### Profile of typical vehicles at each crash severity

```{r}
#change make to keep makes where n > 1000 and change all others to "other"
cars <- crashes_and_vehicles %>% 
  mutate(MAKE_2 = case_when(
    MAKE == "GENERAL MOTORS CORP." ~ "GMC",
    MAKE == "GENERAL MOTORS CORPORATION (GMC)" ~ "GMC",
    MAKE == "HONDA" ~ "HONDA",
    MAKE == "HYUNDAI" ~ "HYUNDAI",
    MAKE == "INFINITI" ~ "INFINITI",
    MAKE == "ISUZU" ~ "ISUZU",
    MAKE == "JAGUAR" ~ "JAGUAR",
    MAKE == "JEEP" ~ "JEEP",
    MAKE == "KIA" ~ "KIA",
    MAKE == "KIA MOTORS CORP" ~ "KIA",
    MAKE == "LAND ROVER" ~ "LAND ROVER",
    MAKE == "LEXUS" ~ "LEXUS",
    MAKE == "LINCOLN" ~ "LINCOLN",
    MAKE == "LINCOLN-CONTINENTAL" ~ "LINCOLN",
    MAKE == "MACK TRUCKS, INC." ~ "MACK TRUCK",
    MAKE == "MAZDA" ~ "MAZDA",
    MAKE == "MERCEDES-BENZ" ~ "MERCEDES",
    MAKE == "MERCURY" ~ "MERCURY",
    MAKE == "MINI" ~ "MINI COOPER",
    MAKE == "MITSUBISHI" ~ "MITSUBISHI",
    MAKE == "NEW FLYER" ~ "NEW FLYER",
    MAKE == "NISSAN" ~ "NISSAN",
    MAKE == "OLDSMOBILE" ~ "OLDSMOBILE",
    MAKE == "PONTIAC" ~ "PONTIAC",
    MAKE == "PONTIAC (CANADIAN)" ~ "PONTIAC",
    MAKE == "PORSCHE" ~ "PORSCHE",
    MAKE == "SATURN" ~ "SATURN",
    MAKE == "SCION" ~ "SCION",
    MAKE == "SUBARU" ~ "SUBARU",
    MAKE == "SUZUKI" ~ "SUZUKI",
    MAKE == "TESLA" ~ "TESLA",
    MAKE == "TESLA MOTORS" ~ "TESLA",
    MAKE == "TOYOTA" ~ "TOYOTA",
    MAKE == "TOYOTA MOTOR COMPANY, LTD." ~ "TOYOTA",
    MAKE == "VOLKSWAGEN" ~ "VOLKSWAGEN",
    MAKE == "VOLVO" ~ "VOLVO",
    TRUE ~ "Other"
  )) %>% 
  mutate(WHERE_FROM = case_when(
    LIC_PLATE_STATE == "IL" ~ "Local",
    TRUE ~ "Other"
  )) %>%  #car from IL or other
  filter(MAKE_2 != "Other") #drop other because I want info on the cars I've actually heard of

#find profile of most frequent vehicles in fatal crashes
fatal_cars <- cars %>% 
  filter(MOST_SEVERE_INJURY == "FATAL")

fatal_car_profile <- fatal_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(fatal_cars)) 

fatal_car_profile[which.max(fatal_car_profile$proportion),] 
#local Toyotas make up the highest proportion of fatal car crashes among major car brands

#find profile of most frequent vehicles in  crashes
fatal_cars <- cars %>% 
  filter(MOST_SEVERE_INJURY == "FATAL")

fatal_car_profile <- fatal_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(fatal_cars)) 

fatal_car_profile[which.max(fatal_car_profile$proportion),] 
#local Toyotas make up the highest proportion of fatal car crashes among major car brands with 15.6%

#find profile of most frequent cars in incapacitating crashes
inc_cars <- cars %>% 
  filter(MOST_SEVERE_INJURY == "INCAPACITATING INJURY")

inc_car_profile <- inc_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(inc_cars)) 

inc_car_profile[which.max(inc_car_profile$proportion),] #local Toyotas are involved in 16.3% of incapacitating crashes in Chicago

#find profile of most frequent cars in non-incapacitating injury crashes
noninc_cars <- cars %>% 
  filter(MOST_SEVERE_INJURY == "NONINCAPACITATING INJURY")

noninc_car_profile <- noninc_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(noninc_cars)) 

noninc_car_profile[which.max(noninc_car_profile$proportion),] #local Toyotas are involved in 17% of non-incapacitating crashes in Chicago

#find profile of most frequent cars in "injuries reported but not evident" crashes
ne_cars <- cars %>% 
  filter(MOST_SEVERE_INJURY == "REPORTED, NOT EVIDENT")

ne_car_profile <- ne_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(ne_cars)) 

ne_car_profile[which.max(ne_car_profile$proportion),] #local Toyotas are involved in 18.2% of "injuries reported but not evident" crashes in Chicago

#find profile of most frequent cars in "no injury" crashes
noinj_cars <- cars %>% 
  filter(MOST_SEVERE_INJURY == "NO INDICATION OF INJURY")

noinj_car_profile <- noinj_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(noinj_cars)) 

noinj_car_profile[which.max(noinj_car_profile$proportion),] #local Toyotas are involved in 18.8% of crashes in Chicago with no injuries
```

Profile of typical drivers at each crash cause

Profile of typical vehicles at each crash cause

Where in the city are crashes of different severities and causes happening?

Do crashes involving cell phone use generally have more injuries, more damage, etc. than crashes not involving cell phone use?

What is the profile of the typical driver and vehicle involved in crashes where BAC \> 0?
