---
title: "DSCI 401 Project"
author: "Rachel Jordan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#load data
crashes <- read.csv("Crashes.csv")
people <- read.csv("People.csv")
vehicles <- read.csv("Vehicles.csv")

#load tidyverse
require(tidyverse)

#pare down variables to only the ones I want
crashes <- crashes %>% 
  select(CRASH_RECORD_ID,CRASH_DATE,POSTED_SPEED_LIMIT,WEATHER_CONDITION,
         LIGHTING_CONDITION,CRASH_TYPE,DAMAGE,PRIM_CONTRIBUTORY_CAUSE,
         SEC_CONTRIBUTORY_CAUSE,STREET_NO,STREET_DIRECTION,STREET_NAME,
         MOST_SEVERE_INJURY,INJURIES_TOTAL,INJURIES_FATAL,
         INJURIES_INCAPACITATING,INJURIES_NON_INCAPACITATING,
         INJURIES_REPORTED_NOT_EVIDENT, INJURIES_NO_INDICATION,
         INJURIES_UNKNOWN,CRASH_HOUR,CRASH_DAY_OF_WEEK,CRASH_MONTH,
         LATITUDE,LONGITUDE,LOCATION)

people <- people %>% 
  select(PERSON_ID,PERSON_TYPE,CRASH_RECORD_ID,SEX,AGE,DRIVERS_LICENSE_STATE,
         EJECTION,INJURY_CLASSIFICATION,DRIVER_ACTION,DRIVER_VISION,
         BAC_RESULT,BAC_RESULT.VALUE,CELL_PHONE_USE)

vehicles <- vehicles %>% 
  select(CRASH_RECORD_ID,MAKE,MODEL,LIC_PLATE_STATE,VEHICLE_YEAR)

#join data sets
crashes_and_people <- people %>% 
  left_join(crashes,by=c("CRASH_RECORD_ID" = "CRASH_RECORD_ID"))
#each record is a person

crashes_and_vehicles <- vehicles %>% 
  left_join(crashes,by=c("CRASH_RECORD_ID" = "CRASH_RECORD_ID"))
#each record is a vehicle

```

#### Profile of typical drivers at each crash severity

```{r}
#filter for only driver information
drivers <- crashes_and_people %>% 
  filter(PERSON_TYPE == "DRIVER")

#filter out blank gender (97 records are blank for gender)
drivers <- drivers %>% 
  filter(SEX == "F" | SEX == "M" | SEX == "X")
#unclear if X is non-binary, NA, or a mix

#change drivers license state to local or other
drivers <- drivers %>% 
  drop_na(DRIVERS_LICENSE_STATE) %>% 
  mutate(WHERE_FROM = case_when(
    DRIVERS_LICENSE_STATE == "IL" ~ "Local",
    TRUE ~ "Other"
  ))

#find profile of most frequent drivers in fatal crashes
fatal_drivers <- drivers %>% 
  filter(MOST_SEVERE_INJURY == "FATAL")

fatal_driver_profile <- fatal_drivers %>% 
  drop_na(AGE) %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(fatal_drivers), mean_age = mean(AGE)) 

fatal_driver_profile[which.max(fatal_driver_profile$proportion),] #male locals around age 39 make up 53% of fatal crashes in Chicago

#find profile of most frequent drivers in incapacitating crashes
inc_drivers <- drivers %>% 
  filter(MOST_SEVERE_INJURY == "INCAPACITATING INJURY")

inc_driver_profile <- inc_drivers %>% 
  drop_na(AGE) %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(inc_drivers), mean_age = mean(AGE)) 

inc_driver_profile[which.max(inc_driver_profile$proportion),] #male locals around age 40 make up 46.3% of incapacitating crashes in Chicago

#find profile of most frequent drivers in non-incapacitating injury crashes
noninc_drivers <- drivers %>% 
  filter(MOST_SEVERE_INJURY == "NONINCAPACITATING INJURY")

noninc_driver_profile <- noninc_drivers %>% 
  drop_na(AGE) %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(noninc_drivers), mean_age = mean(AGE)) 

noninc_driver_profile[which.max(noninc_driver_profile$proportion),] #male locals around age 40 make up 43.7% of non-incapacitating injury crashes in Chicago

#find profile of most frequent drivers in "injuries reported but not evident" crashes
ne_drivers <- drivers %>% 
  filter(MOST_SEVERE_INJURY == "REPORTED, NOT EVIDENT")

ne_driver_profile <- ne_drivers %>% 
  drop_na(AGE) %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(ne_drivers), mean_age = mean(AGE)) 

ne_driver_profile[which.max(ne_driver_profile$proportion),] #male locals around age 41 make up 42% of "injuries reported but not evident" crashes in Chicago

#find profile of most frequent drivers in "no injury" crashes
noinj_drivers <- drivers %>% 
  filter(MOST_SEVERE_INJURY == "NO INDICATION OF INJURY")

noinj_driver_profile <- noinj_drivers %>% 
  drop_na(AGE) %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(noinj_drivers), mean_age = mean(AGE)) 

noinj_driver_profile[which.max(noinj_driver_profile$proportion),] #male locals around age 41 make up 38.4% of crashes in Chicago with no injuries
```

#### Profile of typical vehicles at each crash severity

```{r}
#change make to keep makes where n > 1000 and change all others to "other"
cars <- crashes_and_vehicles %>% 
  mutate(MAKE_2 = case_when(
    MAKE == "GENERAL MOTORS CORP." ~ "GMC",
    MAKE == "GENERAL MOTORS CORPORATION (GMC)" ~ "GMC",
    MAKE == "HONDA" ~ "HONDA",
    MAKE == "HYUNDAI" ~ "HYUNDAI",
    MAKE == "INFINITI" ~ "INFINITI",
    MAKE == "ISUZU" ~ "ISUZU",
    MAKE == "JAGUAR" ~ "JAGUAR",
    MAKE == "JEEP" ~ "JEEP",
    MAKE == "KIA" ~ "KIA",
    MAKE == "KIA MOTORS CORP" ~ "KIA",
    MAKE == "LAND ROVER" ~ "LAND ROVER",
    MAKE == "LEXUS" ~ "LEXUS",
    MAKE == "LINCOLN" ~ "LINCOLN",
    MAKE == "LINCOLN-CONTINENTAL" ~ "LINCOLN",
    MAKE == "MACK TRUCKS, INC." ~ "MACK TRUCK",
    MAKE == "MAZDA" ~ "MAZDA",
    MAKE == "MERCEDES-BENZ" ~ "MERCEDES",
    MAKE == "MERCURY" ~ "MERCURY",
    MAKE == "MINI" ~ "MINI COOPER",
    MAKE == "MITSUBISHI" ~ "MITSUBISHI",
    MAKE == "NEW FLYER" ~ "NEW FLYER",
    MAKE == "NISSAN" ~ "NISSAN",
    MAKE == "OLDSMOBILE" ~ "OLDSMOBILE",
    MAKE == "PONTIAC" ~ "PONTIAC",
    MAKE == "PONTIAC (CANADIAN)" ~ "PONTIAC",
    MAKE == "PORSCHE" ~ "PORSCHE",
    MAKE == "SATURN" ~ "SATURN",
    MAKE == "SCION" ~ "SCION",
    MAKE == "SUBARU" ~ "SUBARU",
    MAKE == "SUZUKI" ~ "SUZUKI",
    MAKE == "TESLA" ~ "TESLA",
    MAKE == "TESLA MOTORS" ~ "TESLA",
    MAKE == "TOYOTA" ~ "TOYOTA",
    MAKE == "TOYOTA MOTOR COMPANY, LTD." ~ "TOYOTA",
    MAKE == "VOLKSWAGEN" ~ "VOLKSWAGEN",
    MAKE == "VOLVO" ~ "VOLVO",
    TRUE ~ "Other"
  )) %>% 
  mutate(WHERE_FROM = case_when(
    LIC_PLATE_STATE == "IL" ~ "Local",
    TRUE ~ "Other"
  )) %>%  #car from IL or other
  filter(MAKE_2 != "Other") #drop other because I want info on the cars I've actually heard of

#find profile of most frequent vehicles in fatal crashes
fatal_cars <- cars %>% 
  filter(MOST_SEVERE_INJURY == "FATAL")

fatal_car_profile <- fatal_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(fatal_cars)) 

fatal_car_profile[which.max(fatal_car_profile$proportion),] 
#local Toyotas make up the highest proportion of fatal car crashes among major car brands at 18.2%

#find profile of most frequent vehicles in  crashes
fatal_cars <- cars %>% 
  filter(MOST_SEVERE_INJURY == "FATAL")

fatal_car_profile <- fatal_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(fatal_cars)) 

fatal_car_profile[which.max(fatal_car_profile$proportion),] 
#local Toyotas make up the highest proportion of fatal car crashes among major car brands with 15.6%

#find profile of most frequent cars in incapacitating crashes
inc_cars <- cars %>% 
  filter(MOST_SEVERE_INJURY == "INCAPACITATING INJURY")

inc_car_profile <- inc_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(inc_cars)) 

inc_car_profile[which.max(inc_car_profile$proportion),] #local Toyotas are involved in 16.3% of incapacitating crashes in Chicago

#find profile of most frequent cars in non-incapacitating injury crashes
noninc_cars <- cars %>% 
  filter(MOST_SEVERE_INJURY == "NONINCAPACITATING INJURY")

noninc_car_profile <- noninc_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(noninc_cars)) 

noninc_car_profile[which.max(noninc_car_profile$proportion),] #local Toyotas are involved in 17% of non-incapacitating crashes in Chicago

#find profile of most frequent cars in "injuries reported but not evident" crashes
ne_cars <- cars %>% 
  filter(MOST_SEVERE_INJURY == "REPORTED, NOT EVIDENT")

ne_car_profile <- ne_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(ne_cars)) 

ne_car_profile[which.max(ne_car_profile$proportion),] #local Toyotas are involved in 18.2% of "injuries reported but not evident" crashes in Chicago

#find profile of most frequent cars in "no injury" crashes
noinj_cars <- cars %>% 
  filter(MOST_SEVERE_INJURY == "NO INDICATION OF INJURY")

noinj_car_profile <- noinj_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(noinj_cars)) 

noinj_car_profile[which.max(noinj_car_profile$proportion),] #local Toyotas are involved in 18.8% of crashes in Chicago with no injuries

```

#### Profile of typical drivers at each crash cause (5 most common crash causes)

```{r}
#determine 5 most common crash causes besides #1 and #6 which are unknown and not applicable
drivers %>% 
  group_by(PRIM_CONTRIBUTORY_CAUSE) %>% 
  summarize(n = n()) %>% 
  arrange(desc(n)) %>% 
  head(7)

#FAILING TO YIELD RIGHT-OF-WAY
#FOLLOWING TOO CLOSELY
#IMPROPER OVERTAKING/PASSING
#FAILING TO REDUCE SPEED TO AVOID CRASH
#IMPROPER LANE USAGE

#find profile of most frequent drivers in failing-to-yield crashes
fty_drivers <- drivers %>% 
  filter(PRIM_CONTRIBUTORY_CAUSE == "FAILING TO YIELD RIGHT-OF-WAY")

fty_driver_profile <- fty_drivers %>% 
  drop_na(AGE) %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(fty_drivers), mean_age = mean(AGE)) 

fty_driver_profile[which.max(fty_driver_profile$proportion),] #male locals around age 41 make up 44.4% of crashes in Chicago where the driver failed to yield

#find profile of most frequent drivers in "following too closely" crashes
ftc_drivers <- drivers %>% 
  filter(PRIM_CONTRIBUTORY_CAUSE == "FOLLOWING TOO CLOSELY")

ftc_driver_profile <- ftc_drivers %>% 
  drop_na(AGE) %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(ftc_drivers), mean_age = mean(AGE)) 

ftc_driver_profile[which.max(ftc_driver_profile$proportion),] #male locals around age 41 make up 42% of crashes in Chicago where the driver was following too closely

#find profile of most frequent drivers in "improper overtaking/passing crashes
iop_drivers <- drivers %>% 
  filter(PRIM_CONTRIBUTORY_CAUSE == "IMPROPER OVERTAKING/PASSING")

iop_driver_profile <- iop_drivers %>% 
  drop_na(AGE) %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(iop_drivers), mean_age = mean(AGE)) 

iop_driver_profile[which.max(iop_driver_profile$proportion),] #male locals around age 42 make up 40% of crashes in Chicago where the driver improperly overtook/passed someone

#find profile of most frequent drivers in "FAILING TO REDUCE SPEED TO AVOID CRASH" crashes
ftrs_drivers <- drivers %>% 
  filter(PRIM_CONTRIBUTORY_CAUSE == "FAILING TO REDUCE SPEED TO AVOID CRASH")

ftrs_driver_profile <- ftrs_drivers %>% 
  drop_na(AGE) %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(ftrs_drivers), mean_age = mean(AGE)) 

ftrs_driver_profile[which.max(ftrs_driver_profile$proportion),] #male locals around age 40 make up 44.1% of crashes in Chicago where the driver failed to yield

#find profile of most frequent drivers in "improper lane usage" crashes
ilu_drivers <- drivers %>% 
  filter(PRIM_CONTRIBUTORY_CAUSE == "IMPROPER LANE USAGE")

ilu_driver_profile <- ilu_drivers %>% 
  drop_na(AGE) %>% 
  group_by(SEX,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(ilu_drivers), mean_age = mean(AGE)) 

ilu_driver_profile[which.max(ilu_driver_profile$proportion),] #male locals around age 42 make up 42.6% of crashes in Chicago where the driver used lane improperly
```

#### Profile of typical vehicles at each crash cause

```{r}
#find profile of most frequent vehicles in "FAILING TO YIELD RIGHT-OF-WAY" crashes
fty_cars <- cars %>% 
  filter(PRIM_CONTRIBUTORY_CAUSE == "FAILING TO YIELD RIGHT-OF-WAY")

fty_car_profile <- fty_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(fty_cars)) 

fty_car_profile[which.max(fty_car_profile$proportion),] 
#local Toyotas make up the highest proportion of "FAILING TO YIELD RIGHT-OF-WAY" car crashes among major car brands at 20%

#find profile of most frequent vehicles in "FOLLOWING TOO CLOSELY" crashes
ftc_cars <- cars %>% 
  filter(PRIM_CONTRIBUTORY_CAUSE == "FOLLOWING TOO CLOSELY")

ftc_car_profile <- ftc_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(ftc_cars)) 

ftc_car_profile[which.max(ftc_car_profile$proportion),] 
#local Toyotas make up the highest proportion of "FAILING TO YIELD RIGHT-OF-WAY" car crashes among major car brands at 19.4%

#find profile of most frequent vehicles in "FOLLOWING TOO CLOSELY" crashes
ftc_cars <- cars %>% 
  filter(PRIM_CONTRIBUTORY_CAUSE == "FOLLOWING TOO CLOSELY")

ftc_car_profile <- ftc_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(ftc_cars)) 

ftc_car_profile[which.max(ftc_car_profile$proportion),] 
#local Toyotas make up the highest proportion of "FOLLOWING TOO CLOSELY" car crashes among major car brands at 19.4%

#find profile of most frequent vehicles in "IMPROPER OVERTAKING/PASSING" crashes
iop_cars <- cars %>% 
  filter(PRIM_CONTRIBUTORY_CAUSE == "IMPROPER OVERTAKING/PASSING")

iop_car_profile <- iop_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(iop_cars)) 

iop_car_profile[which.max(iop_car_profile$proportion),] 
#local Toyotas make up the highest proportion of "IMPROPER OVERTAKING/PASSING" car crashes among major car brands at 19.6%

#find profile of most frequent vehicles in "IMPROPER OVERTAKING/PASSING" crashes
iop_cars <- cars %>% 
  filter(PRIM_CONTRIBUTORY_CAUSE == "IMPROPER OVERTAKING/PASSING")

iop_car_profile <- iop_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(iop_cars)) 

iop_car_profile[which.max(iop_car_profile$proportion),] 
#local Toyotas make up the highest proportion of "IMPROPER OVERTAKING/PASSING" car crashes among major car brands at 19.6%

#find profile of most frequent vehicles in "FAILING TO REDUCE SPEED TO AVOID CRASH" crashes
ftrs_cars <- cars %>% 
  filter(PRIM_CONTRIBUTORY_CAUSE == "FAILING TO REDUCE SPEED TO AVOID CRASH")

ftrs_car_profile <- ftrs_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(ftrs_cars)) 

ftrs_car_profile[which.max(ftrs_car_profile$proportion),] 
#local Toyotas make up the highest proportion of "FAILING TO REDUCE SPEED TO AVOID CRASH" car crashes among major car brands at 17.1%

#find profile of most frequent vehicles in "IMPROPER LANE USAGE" crashes
ilu_cars <- cars %>% 
  filter(PRIM_CONTRIBUTORY_CAUSE == "IMPROPER LANE USAGE")

ilu_car_profile <- ilu_cars %>% 
  group_by(MAKE_2,WHERE_FROM) %>% 
  summarize(proportion = n()/nrow(ilu_cars)) 

ilu_car_profile[which.max(ilu_car_profile$proportion),] 
#local Toyotas make up the highest proportion of "IMPROPER LANE" car crashes among major car brands at 20.4%
```

#### Where in the city are crashes of different severities and causes happening?

```{r}
#find injury locations
injury_locations <- crashes %>% 
  drop_na(LATITUDE,LONGITUDE) %>% 
  filter(MOST_SEVERE_INJURY == "FATAL" | MOST_SEVERE_INJURY == "INCAPACITATING INJURY"| MOST_SEVERE_INJURY == "NONINCAPACITATING INJURY" | MOST_SEVERE_INJURY == "REPORTED, NOT EVIDENT") %>% 
  filter(LONGITUDE != 0)

#find cause locations
cause_locations <- crashes %>% 
  filter(PRIM_CONTRIBUTORY_CAUSE == "FAILING TO YIELD RIGHT-OF-WAY" | PRIM_CONTRIBUTORY_CAUSE == "FOLLOWING TOO CLOSELY" | PRIM_CONTRIBUTORY_CAUSE == "IMPROPER OVERTAKING/PASSING" | PRIM_CONTRIBUTORY_CAUSE == "FAILING TO REDUCE SPEED TO AVOID CRASH" | PRIM_CONTRIBUTORY_CAUSE == "IMPROPER LANE USAGE") %>% 
  drop_na(LATITUDE, LONGITUDE) %>% 
  filter(LATITUDE != 0)

#map
require(rgdal)

#create shapefile
ogrInfo("Boundaries - Community Areas (current).geojson")
shapefile <- readOGR(dsn="Boundaries - Community Areas (current).geojson", p4s="+proj=tmerc +ellps=WGS84")
shapefile_df <- fortify(shapefile)

#map of crashes by injury severity
ggplot() + geom_polygon(shapefile_df, mapping=aes(x=long, y=lat, group=group)) + geom_path()  + geom_point(injury_locations[sample(nrow(injury_locations),nrow(injury_locations)*0.05,replace=F),],mapping=aes(x=LONGITUDE,y=LATITUDE, color=MOST_SEVERE_INJURY),size=1.0) + ggtitle("Crash Locations") #removed no injury crashes bc they were obscuring everything else. plotted 5% of the data as selected by random sampling because of the sheer amount of observations -- the map was unintelligible with the full data set. less activity in far south side and north west side, more activity in an east/west band around lakeview/lincoln park and west of lakeview. less activity around ~little village, englewood. less fatal injuries seem to happen in the loop.

#map of crashes by cause
ggplot() + geom_polygon(shapefile_df, mapping=aes(x=long, y=lat, group=group)) + geom_path()  + geom_point(cause_locations[sample(nrow(cause_locations),nrow(cause_locations)*0.05,replace=F),],mapping=aes(x=LONGITUDE,y=LATITUDE, color=PRIM_CONTRIBUTORY_CAUSE),size=1.0) + ggtitle("Crash Locations") #removed no injury crashes bc they were obscuring everything else. plotted 5% of the data as selected by random sampling because of the sheer amount of observations -- the map was unintelligible with the full data set. no patterns indicated by the map except less activity in on the south side, especially the far south side. maybe slightly more improper lane usage and improper overtaking/passing in the loop. 
```

#### Do crashes involving cell phone use generally have more injuries, more damage, etc. than crashes not involving cell phone use?

```{r}
#create data set with cell phone use
cell_phone_use <- crashes_and_people %>% 
  mutate(cell_phone = case_when(
    PRIM_CONTRIBUTORY_CAUSE == "TEXTING" ~ "YES",
    PRIM_CONTRIBUTORY_CAUSE == "CELL PHONE USE OTHER THAN TEXTING" ~ "YES",
    TRUE ~ "NO"
  )) %>% 
  select(cell_phone,DAMAGE,INJURIES_TOTAL)

yes_cell <- cell_phone_use %>% 
  drop_na(DAMAGE) %>% 
  filter(cell_phone == "YES")

no_cell <- cell_phone_use %>% 
  drop_na(DAMAGE) %>% 
  filter(cell_phone == "NO")

cell_phone_use %>% 
  drop_na(DAMAGE) %>% 
  group_by(cell_phone,DAMAGE) %>% 
  summarize(proportion_yes = n()/nrow(yes_cell), proportion_no = n()/nrow(no_cell), mean_injuries = mean(INJURIES_TOTAL)) #ignore proportion yes for cell phone no and proportion no for cell phone yes
#looks like there is slightly more severe damage when there is cell phone use involved, but we don't know if it's STATISTICALLY significant
#there do appear to be more injuries with cell phone use but again not sure if STATISTICALLY significant
```

#### What is the profile of the typical driver and vehicle involved in crashes where BAC \> 0?
